Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/intro.tex
uncomment splitting section

\begin{comment}
\section{Introduction}

Such computations, where we apply a multivariate function to arguments
where one of the argument remains the same are common.  For example,
many algorithm-design techniques revolve around the idea of finding
the computation in $f$ that depends only on $x$, and performing them
once and for all to use.  Similarly, crucial compiler optimization
techniques such as loop hoisting and common subexpression elimination
identify common computations across all invocations of $f$ and perform
them once and for all at the beginning of the computation.  A
particular application area where such algorithm-design and code
optimization techniques are widely applied include real-time computer
graphics.  There, to achieve high-performance, programmers manually
separate code into distinct {\em passes} based on their required
frequency of execution.  For example, graphics computations are often
expressed as a sequence of passes where each pass may perform a
computation in increasingly finer grain, e.g., per object, per region
of the screen, and per pixel, where later computations use the results
from the prior computations.

%
While performant, this programming style of programming with manually
separated passes results in complex code where invariants must hold
across different passes, and local changes for one pass may require
changes to other passes to ``plumb'' inputs from all other
frequencies\,\cite{Foley:2011}.
%
In other words, manual pass separation breaks key programming
abstractions such as composition and modularity that we hold dear.
%
Graphics researchers therefore have suggested writing graphics
programs in explicitly staged programming
languages\,\cite{Proudfoot:2001,Foley:2011,He:2014}, deferring the
tedious task of separating the code into passes as needed for
performance to the compiler. However, all prior efforts have been
limited to simple languages that do not support recursion or
first-class functions.

Motivated by the crucial need for improved abstractions for performant
multi-stage programs, we investigate in this paper, the problem of
automatically splitting a higher-order program into a collection of
programs each of which gets executed in a single pass.
%
This is not a new problem.
%
Common theme in this problem domain is what in the 1980s J{\o}rring
and Scherlis called {\em frequency reduction} or {\em
  precomputation}~\cite{JS86-staging}.  The idea behind frequency
reduction is to identify computations that are performed multiple
times and hoist them so that they can be performed once and used later
as needed.  The idea behind precomputation is to identify computations
that can be performed earlier---for example, at compile time if their
inputs are known statically---and perform them at that earlier time.

There has been much work on language-based techniques for performing
frequency reduction and precomputation.  Broadly speaking, these
techniques can be categorized into
\begin{itemize}
\item partial evaluation,
\item metaprogramming, and
\item pass separation or data specialization.
\end{itemize}

Partial evaluation achieves precomputation by distinguishing between
{\em static} and {\em dynamic} stages, and performing all static
computations first, followed by dynamic ones.  For example, suppose
that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a statically
known $x = [1,2, \ldots, n]$.  We can derive a specialization of $f$
at $[1,2, \ldots, n]$, written $f_{[1,2, \ldots, n]}$, such that
$f(x,y) = f_{[1,2, \ldots, n]}(y)$ and specialize $F$ for $x=[~]$ as
\begin{lstlisting}
fun @$F_{[1,2, \ldots, n]}(y_0, y_1 \ldots y_{m-1})$@ = 
  @$(f_{[1,2, \ldots, n]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
\end{lstlisting}
%
Such a partial evaluation typically proceed by performing a
binding-time analysis to determine the static and dynamic parts of $f$
and $F$ and specializes them based on the results of the binding-time
analysis.

A closely related technique, {\em metaprogramming}, enables the
programmer to write a program that takes the static values and
generates a specialized program for those values.  Metaprogramming
gives the programmer relatively fine-grained control over partial
evaluation by requiring staging annotations in the program.  Such
staging annotations make explicit the stage of each computation and
can overcome some of the intricacies of binding-time analysis but this
comes at the cost of more complex reasoning and some loss of
automation (partial evaluation can in principle be applied fully
automatically without relying on any annotations).

The third and final approach, called {\em pass separation} or {\em
  data specialization}, assumes much less information: instead of
requiring the static data, it requires just the knowledge of which
computations are static and dynamic and uses this knowledge to
generate two stage computations. 
%
Applied to our example, pass separation would {\em split} the function
$f$ into $f_s$ and $f_d$, corresponding to the static and dynamic
passes, such that $f(x,y) = f_d(f_s(x),y)$.  We can then rewrite our
example as
%
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1})$@ = 
let @$fx = f_s(x)$@
in @$(f_d(fx, y_0);  f_d(fx, y_1)  \ldots f_d(fx, y_{m-1}))$@.
\end{lstlisting}
We refer to such a program as a {\em multi-pass} program.

The key difference with partial evaluated program and the stratified
program is that the latter has lost no generality: it works with all
arguments, whereas the partially evaluated program is specialized for
a particular value of $x$ ($x = [1,2, \ldots, n]$ in the example).
Thus, the pass separation can be applied without leading to any loss
of generality.  In fact, this is quite natural---in the terminology of
J{\o}rring and Scherlis, pass separation is a frequency reduction,
whereas as partial evaluation and metaprogramming are precomputation
techniques. Consequently, if desired, partial evaluation can be
orthogonally applied to a multi-pass program.
%
The crucial benefit of pass-separation is efficiency: separating the
passes in the program factors out frequently used the parts of
computation, computes them first, and then reuse them later instead of
computing them again. This is precisely the reason for why it is
employed broadly in graphics software even though it results in
code that is difficult to design, implement, and maintain.

Thus, we are interested in the question of whether it would be
possible to obtain the best of the both worlds, by automatically
pass-separating programs.  Such automatic stratification has been
shown to be challenging. The first successful approach that we know of
is by Knoblock and Ruf\,\cite{knoblock96}, which, like our work, was
inspired by computer graphics applications.  While practically
effective, Knoblock and Ruf consider only a very simple language and
are not able to stratify recursive programs.  Recent
efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} ....  but still
cannot stratify recursive programs.

In this paper we make significant progress on the problem of automatic
stratification of programs expressed in a rich higher-order functional
language.  As our starting point, we take the modal language \lang\
with explicit staging in the style of Davies~\cite{davies96}, where
the ``circle'' modality denotes computation in a later stage, and
present a staged operational semantics similar for \lang\
(\ref{sec:stagedsemantics}).  We then present a splitting algorithm
that stratifies a \lang\ program into a two mono-staged programs, one
for the first stage and for the second stage (\ref{sec:splitting}).
Since each program is monostaged, they are expressed in conventional
functional languages.

@@

We start in \ref{sec:example} with an extended example of splitting
\texttt{quickselect}.
In \ref{sec:semantics}, we describe our staged language \lang, including its
type system and operational semantics.
In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
algorithm for performing data specialization.
Finally, in \ref{sec:examples}, we show how our stage-splitting algorithm
transforms a variety of other programs.
\end{comment}

>>>>

@@

We start in \ref{sec:example} with an extended example of splitting
\texttt{quickselect}.
In \ref{sec:semantics}, we describe our staged language \lang, including its
type system and operational semantics.
In \ref{sec:splitting,sec:implementation}, we describe our stage-splitting
algorithm for performing data specialization.
Finally, in \ref{sec:examples}, we show how our stage-splitting algorithm
transforms a variety of other programs.
